---
title: "Multisite Analysis of Neurologic Disease and HMPV"
subtitle: "Expanding the Single-Site Pittsburgh Findings (2024)"
author:
  name: "Ray Pomponio, for Dr. Anna Wang-Erickson"
  affiliation: "Children's Hospital of Pittsburgh"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
    embed-resources: true
    fig-cap-location: top
abstract-title: "Study Summary"
abstract: |
  We expanded the single-site Pittsburgh analysis (2024) to the full multisite NVSN cohort to investigate whether children with neurologic/neuromuscular disease experience disproportionately higher risk of human metapneumovirus (HMPV) infection and disease severity. Previous single-site findings showed neurologic disease was associated with 1.75× higher odds of HMPV infection (95% CI: 1.19, 2.56) and 4.80× higher odds of elevated severity (95% CI: 2.01, 11.47). This multisite analysis (136,156 cases across 7 sites, 2016–2026) adjusts for demographics, comorbidities, and geographic variation to assess whether neurologic conditions confer disproportionate risk in a larger, more generalizable cohort.
---

```{r}
#| warning: false
#| message: false

# --- LOAD PACKAGES ---
library(Hmisc)
library(data.table)
library(rms)
library(mice)
library(gtsummary)
library(ggplot2)
library(gt)
library(dplyr)
theme_set(theme_classic(base_size = 12))
options(prType = "html")

# --- CONSTANTS & RANDOM SEED ---
N.IMPUTE <- 5
SEED.MULTISITE <- 7429
set.seed(SEED.MULTISITE)

# --- VARIABLE LABELS (Semantic meanings for tables) ---
var.labels <- list(
  c_agegroup = "Age Group",
  sexch = "Sex",
  c_race_int = "Race/Ethnicity",
  scrinsurance = "Insurance",
  c_ariyear = "Study Year",
  studysite = "Study Site",
  c_uresp = "Chronic Respiratory Condition",
  c_ucardiovasc = "Cardiovascular Disease",
  c_uoncolimmu = "Immunosuppressive Condition",
  c_uprem = "Prematurity",
  c_unuerologic = "Neurologic/Neuromuscular Disease",
  c_hmpvcoinfected = "Viral Co-infection",
  c_hmpv.pos = "HMPV Status",
  c_severity = "Illness Severity"
)
```

# Introduction

This report presents a multisite analysis extending our 2024 single-site (Pittsburgh) findings on the association between neurologic/neuromuscular disease and human metapneumovirus (HMPV) infection and disease severity. We leverage 136,156 pediatric acute respiratory infection (ARI) cases across 7 NVSN sites over 10+ years (2016–2026) to investigate whether neurologic conditions represent a consistent, significant risk factor for HMPV infection and severe outcomes in a larger, more representative cohort.

## Previous Findings (Pittsburgh, 2024)

In our single-site analysis, we found:
- **HMPV infection risk:** Children with neurologic/neuromuscular disease had 1.75× higher odds of HMPV infection (95% CI: 1.19, 2.56)
- **HMPV disease severity:** Among HMPV-positive cases, neurologic disease was associated with 4.80× higher odds of elevated severity (95% CI: 2.01, 11.47)

## Research Questions

**Primary:** Is neurologic/neuromuscular disease associated with increased HMPV infection risk in a multisite cohort (adjusting for demographics, comorbidities, site)?

**Secondary:** Among HMPV-positive children, is neurologic disease associated with elevated disease severity (5-level ordinal outcome)?

---

# Methods

## Data & Study Population

Data are from the New Vaccine Surveillance Network (NVSN) multisite surveillance network covering pediatric ARI cases from 7 sites (Vanderbilt, Rochester, Cincinnati, Seattle, Houston, Kansas City, Pittsburgh) over study years 2–11 (2016–2026).

**Inclusion criteria:** All pediatric ARI cases with valid HMPV test results (`tmpv` ∈ {0, 1}).

**Exclusion criteria:** HMPV inconclusive or missing results (`tmpv` ∈ {8, 9, NA}).

## Statistical Analysis

### RQ1: HMPV Infection Risk

We assessed factors associated with HMPV infection status using generalized linear models (GLM) with binomial family and logit link. The primary exposure was neurologic/neuromuscular disease (`c_unuerologic`). Adjustment variables included age group, sex, race/ethnicity, insurance, chronic respiratory conditions, cardiovascular disease, immunosuppressive conditions, prematurity, study year, and study site.

Primary analysis used multiple imputation (m=5) for missing covariates, with estimates pooled per Rubin's rules. Unadjusted (univariate) and adjusted (multivariable) odds ratios (ORs) with 95% confidence intervals (CIs) and Wald test p-values are reported.

### RQ2: HMPV Severity Among Positive Cases

We assessed the association between neurologic disease and ordinal illness severity among HMPV-positive cases (n ≈ 6,415) using proportional odds (ordinal logistic) regression. Severity was defined as a 5-level ordinal composite based on clinical setting and supportive care intensity:

- **Level 1:** ED/clinic discharge or outpatient care
- **Level 2:** Hospital admission without supplemental oxygen
- **Level 3:** Hospital admission with supplemental oxygen (any modality: standard, high-flow nasal cannula, blow-by, or CPAP)
- **Level 4:** ICU admission without mechanical ventilation or ECMO
- **Level 5:** ICU admission with mechanical ventilation/ECMO or death

Adjustment variables included age group, sex, race/ethnicity, insurance, comorbidities (respiratory, cardiovascular, immunosuppressive, prematurity), viral co-infections, and study site. We tested the proportional odds assumption using the Brant test and report violations as limitations on model interpretation.

---

# Results

## Phase 1: Data Preparation & Quality Audit

```{r}
#| label: data-load
#| echo: false

dat.raw <- fread(
  "/Users/raypomponio/Repos/NVSN_HMPV_neurology/Data/Pitt_Anna_HMPV_SEP26.csv"
)

cat("=== RAW DATA DIMENSIONS ===\n")
cat("N rows:", nrow(dat.raw), "\n")
cat("N cols:", ncol(dat.raw), "\n\n")

cat("=== HMPV STATUS DISTRIBUTION (tmpv) ===\n")
print(table(dat.raw$tmpv, useNA = "ifany"))
cat("\n")
```

```{r}
#| label: data-prep

# --- STEP 1: DEFINE STUDY POPULATION ---
dat.analyze <- dat.raw[tmpv %in% c(0, 1)]

cat("After restricting to valid HMPV results: N =", nrow(dat.analyze), "\n\n")

# --- STEP 2: RECODE & DERIVE VARIABLES ---
setDT(dat.analyze)

# Age group (4 levels)
dat.analyze[, c_agegroup := fcase(
  c_ageyears < 1, 1,
  c_ageyears < 2, 2,
  c_ageyears < 3, 3,
  c_ageyears >= 3, 4
)]
dat.analyze[, c_agegroup := factor(
  c_agegroup, 4:1, c("3+ yrs", "2-3 yrs", "1-2 yrs", "<1 yr")
)]

# Sex
dat.analyze[, sexch := factor(sexch, 1:2, c("Male", "Female"))]

# Race/ethnicity (White NH as reference)
dat.analyze[, c_race_int := factor(
  c_race_int, c(1, 2, 3, 4), 
  c("WhiteNH", "BlackNH", "OtherNH", "Hispanic")
)]

# Insurance (Private as reference)
dat.analyze[, scrinsurance := factor(
  scrinsurance, c(2, 1, 3, 4),
  c("Private", "Public", "Both", "None")
)]

# Study year (ordered factor)
dat.analyze[, c_ariyear := factor(
  c_ariyear, 2:11, 
  c("Y2 (2016-17)", "Y3 (2017-18)", "Y4 (2018-19)", "Y5 (2019-20)",
    "Y6 (2020-21)", "Y7 (2021-22)", "Y8 (2022-23)", "Y9 (2023-24)",
    "Y10 (2024-25)", "Y11 (2025-26)")
)]

# Study site (Pittsburgh as reference)
dat.analyze[, studysite := factor(
  studysite, c(8, 1, 2, 3, 4, 5, 6),
  c("Pittsburgh", "Vanderbilt", "Rochester", "Cincinnati", 
    "Seattle", "Houston", "Kansas City")
)]

# Comorbidities (binary, 0 = no, 1 = yes)
dat.analyze[, c_uresp := fcase(
  c_uresp == 1, 1, default = 0
)]
dat.analyze[, c_uresp := as.numeric(c_uresp)]

dat.analyze[, c_ucardiovasc := fcase(
  c_ucardiovasc == 1, 1, default = 0
)]
dat.analyze[, c_ucardiovasc := as.numeric(c_ucardiovasc)]

dat.analyze[, c_uoncolimmu := fcase(
  c_uoncolimmu == 1, 1, default = 0
)]
dat.analyze[, c_uoncolimmu := as.numeric(c_uoncolimmu)]

dat.analyze[, c_uprem := fcase(
  c_uprem == 1, 1,
  c_uprem == 0, 0)]
dat.analyze[, c_uprem := as.numeric(c_uprem)]

dat.analyze[, c_unuerologic := fcase(
  c_unuerologic == 1, 1,
  c_unuerologic == 0, 0)]
dat.analyze[, c_unuerologic := as.numeric(c_unuerologic)]

# HMPV binary indicator
dat.analyze[, c_hmpv.pos := fcase(tmpv == 1, 1, default = 0)]

# Co-infection indicator (use existing variable: c_othercoin6_viral)
dat.analyze[, c_hmpvcoinfected := fcase(
  c_hmpv.pos == 1 & c_othercoin6_viral == 1, 1,
  default = 0
)]

# Severity outcome (5-level ordinal for HMPV-positive cases)
# Hierarchical derivation based on setting + support intensity
# Level 1: ED/clinic discharge (c_finalstatus ∈ {2, 3})
# Level 2: Hospital admission without oxygen support
# Level 3: Hospital admission with any oxygen support
# Level 4: ICU without mechanical ventilation/ECMO
# Level 5: ICU with mechanical ventilation/ECMO or death

# Composite oxygen support indicator (any modality)
dat.analyze[, has.oxygen := fcase(
  c_suppoxy == 1, 1,
  c_hfnc == 1, 1,
  c_blowby == 1, 1,
  c_cpap == 1, 1)]

dat.analyze[, c_severity := fcase(
  # Level 5: Death OR mechanical ventilation OR ECMO
  c_died == 1 | c_intubated == 1 | inptEcmo == 1, 5L,
  
  # Level 4: ICU without mechanical support
  inptICU == 1, 4L,
  
  # Level 3: Hospital with oxygen support (any modality)
  c_finalstatus == 1 & has.oxygen == 1, 3L,
  
  # Level 2: Hospital without oxygen support
  c_finalstatus == 1 & (has.oxygen == 0 | inptICU ==0), 2L,
  
  # Level 1: ED/clinic discharge or outpatient
  c_finalstatus %in% c(2, 3, 5), 1L,
  
  default = NA_integer_
)]

dat.analyze[, c_severity := factor(
  c_severity, 1:5,
  c("ED/Clinic Discharge", "Hospital No O₂", "Hospital + O₂",
    "ICU No IMV/ECMO", "ICU with IMV/ECMO or Death")
)]

cat("Variables recoded and derived.\n\n")
```

```{r}
#| label: descriptive-stats

cat("=== DESCRIPTIVE STATISTICS ===\n\n")

cat("N cases (all with valid HMPV result):", nrow(dat.analyze), "\n")
cat("N HMPV-positive:", sum(dat.analyze$c_hmpv.pos), "\n")
cat("N HMPV-negative:", sum(dat.analyze$c_hmpv.pos == 0), "\n\n")

cat("HMPV-positive cases by severity:\n")
print(table(dat.analyze[c_hmpv.pos == 1]$c_severity, useNA = "ifany"))
cat("\n")

cat("Neurologic disease prevalence:\n")
print(table(dat.analyze$c_unuerologic, useNA = "ifany"))
cat("\n")
```

## RQ1: HMPV Infection Risk

**Research Question:** Is neurologic/neuromuscular disease associated with increased HMPV infection risk in a multisite cohort?

**Approach:** We used multiple imputation (m=5) to handle missing covariate data and fitted logistic regression models to estimate univariate and adjusted odds ratios for HMPV infection.

**Results:** See Table 1 below for complete descriptive statistics, univariate ORs, and adjusted ORs with 95% confidence intervals and p-values for all exposures.

```{r}
#| label: mult-imputation
#| echo: false

# --- IDENTIFY COVARIATES FOR IMPUTATION ---
rq1.vars <- c("c_agegroup", "sexch", "c_race_int", "scrinsurance",
              "c_ariyear", "studysite", "c_uresp",
              "c_ucardiovasc", "c_uoncolimmu",
              "c_uprem", "c_unuerologic", "c_hmpv.pos")

dat.rq1 <- dat.analyze[, ..rq1.vars]

cat("=== RQ1 MULTIPLE IMPUTATION (m=5) ===\n")
cat("N cases:", nrow(dat.rq1), "\n")
cat("N covariates:", length(rq1.vars) - 1, "\n\n")

# Convert factors to numeric for mice
mice.data <- dat.rq1[, lapply(.SD, function(x) {
  if (is.factor(x)) as.numeric(x) - 1 else x
})]

# Multiple imputation using mice
imputed.obj <- mice(mice.data, m = N.IMPUTE, method = "pmm", seed = SEED.MULTISITE,
                    printFlag = FALSE)

cat("Imputation completed.\n\n")
```

## RQ2: HMPV Disease Severity

**Research Question:** Among HMPV-positive children, is neurologic/neuromuscular disease associated with elevated disease severity?

**Approach:** We used proportional odds regression with multiple imputation (m=5) to model the 5-level ordinal severity outcome. ORs represent the odds of being in a higher severity category per unit increase in the predictor.

**Results:** See Table 2 below for severity distributions and adjusted ORs with 95% confidence intervals and p-values for all predictors.

```{r}
#| label: rq2-prep
#| echo: false

# Restrict to HMPV-positive cases
dat.hmpv.pos <- dat.analyze[c_hmpv.pos == 1]

cat("=== RQ2 ANALYSIS (HMPV-Positive Cases Only) ===\n")
cat("N cases:", nrow(dat.hmpv.pos), "\n")
cat("Severity distribution:\n")
print(table(dat.hmpv.pos$c_severity, useNA = "ifany"))
cat("\n")

# Exclude only cases with missing severity outcome
rq2.vars <- c("c_severity", "c_agegroup", "sexch", "c_race_int",
              "scrinsurance", "studysite", "c_uresp",
              "c_ucardiovasc", "c_uoncolimmu",
              "c_uprem", "c_unuerologic", "c_hmpvcoinfected")

dat.rq2 <- dat.hmpv.pos[, ..rq2.vars]
dat.rq2 <- dat.rq2[!is.na(c_severity)]

cat("After excluding missing severity: N =", nrow(dat.rq2), "\n\n")

# Convert factors to numeric for mice
mice.data.rq2 <- dat.rq2[, lapply(.SD, function(x) {
  if (is.factor(x)) as.numeric(x) - 1 else x
})]

# Multiple imputation for covariates (m=5)
cat("Running multiple imputation for RQ2 (m=5)...\n")
imputed.obj.rq2 <- mice(mice.data.rq2, m = N.IMPUTE, method = "pmm",
                        seed = SEED.MULTISITE, printFlag = FALSE)
cat("Imputation completed.\n\n")
```

```{r}
#| label: rq2-model
#| echo: false

# Fit proportional odds model across all imputations
cat("=== RQ2 PROPORTIONAL ODDS MODEL (Multiple Imputation) ===\n\n")

rq2.formula <- as.formula(
  "c_severity ~ c_agegroup + sexch + c_race_int + scrinsurance +
   c_uresp + c_ucardiovasc + c_uoncolimmu + c_uprem + c_unuerologic +
   c_hmpvcoinfected + studysite"
)

# Fit model on each imputed dataset and pool
rq2.coefs.all <- list()
rq2.ses.all <- list()

for (i in 1:N.IMPUTE) {
  dat.imp.rq2 <- complete(imputed.obj.rq2, i) |> as.data.frame()
  
  # Restore factor levels
  dat.imp.rq2$c_severity <- factor(
    dat.imp.rq2$c_severity, 0:4,
    c("ED/Clinic Discharge", "Hospital No O₂", "Hospital + O₂",
      "ICU No IMV/ECMO", "ICU with IMV/ECMO or Death")
  )
  dat.imp.rq2$c_agegroup <- factor(dat.imp.rq2$c_agegroup, 3:0,
                                   c("3+ yrs", "2-3 yrs", "1-2 yrs", "<1 yr"))
  dat.imp.rq2$sexch <- factor(dat.imp.rq2$sexch, 0:1, c("Male", "Female"))
  dat.imp.rq2$c_race_int <- factor(dat.imp.rq2$c_race_int, 0:3,
                                   c("WhiteNH", "BlackNH", "OtherNH", "Hispanic"))
  dat.imp.rq2$scrinsurance <- factor(dat.imp.rq2$scrinsurance, 1:4,
                                     c("Private", "Public", "Both", "None"))
  dat.imp.rq2$studysite <- factor(dat.imp.rq2$studysite, c(7, 0:5),
                                  c("Pittsburgh", "Vanderbilt", "Rochester",
                                    "Cincinnati", "Seattle", "Houston", "Kansas City"))
  
  # Fit proportional odds model
  fit.imp <- lrm(rq2.formula, data = dat.imp.rq2, x = TRUE, y = TRUE)
  
  # Extract coefficients and standard errors
  rq2.coefs.all[[i]] <- coef(fit.imp)
  rq2.ses.all[[i]] <- sqrt(diag(vcov(fit.imp)))
}

# Pool results using Rubin's rules
n.coefs.rq2 <- length(rq2.coefs.all[[1]])
rq2.pooled.coefs <- numeric(n.coefs.rq2)
rq2.pooled.ses <- numeric(n.coefs.rq2)
rq2.pooled.pvals <- numeric(n.coefs.rq2)

for (coef.idx in 1:n.coefs.rq2) {
  coef.vec <- sapply(rq2.coefs.all, function(x) x[coef.idx])
  se.vec <- sapply(rq2.ses.all, function(x) x[coef.idx])
  
  mean.coef <- mean(coef.vec)
  within.var <- mean(se.vec^2)
  between.var <- var(coef.vec)
  total.var <- within.var + (1 + 1/N.IMPUTE) * between.var
  pool.se <- sqrt(total.var)
  
  pool.z <- mean.coef / pool.se
  pool.pval <- 2 * (1 - pnorm(abs(pool.z)))
  
  rq2.pooled.coefs[coef.idx] <- mean.coef
  rq2.pooled.ses[coef.idx] <- pool.se
  rq2.pooled.pvals[coef.idx] <- pool.pval
}

names(rq2.pooled.coefs) <- names(rq2.coefs.all[[1]])

cat("Proportional odds model fitted across", N.IMPUTE, "imputed datasets.\n")
cat("Results pooled using Rubin's rules.\n")
cat("Full results presented in Table 2.\n\n")
```

```{r}
#| label: rq2-summary
#| echo: false

cat("=== RQ2 PRIMARY EXPOSURE: NEUROLOGIC DISEASE ===\n\n")

# Extract neurologic disease coefficient
neuro.idx <- grep("c_unuerologic", names(rq2.pooled.coefs))

if (length(neuro.idx) > 0) {
  neuro.coef <- rq2.pooled.coefs[neuro.idx[1]]
  neuro.se <- rq2.pooled.ses[neuro.idx[1]]
  neuro.pval <- rq2.pooled.pvals[neuro.idx[1]]
  
  neuro.or <- exp(neuro.coef)
  neuro.ci.lo <- exp(neuro.coef - 1.96 * neuro.se)
  neuro.ci.hi <- exp(neuro.coef + 1.96 * neuro.se)
  
  cat("Neurologic/Neuromuscular Disease:\n")
  cat(sprintf("  Adjusted OR: %.2f (per severity level increase)\n", neuro.or))
  cat(sprintf("  95%% CI: (%.2f, %.2f)\n", neuro.ci.lo, neuro.ci.hi))
  cat(sprintf("  p-value: %.4f\n\n", neuro.pval))
}

cat("Note: Proportional odds assumption test available in Appendix (Section S2).\n\n")
```

## Table 1: Overall Cohort Characteristics and HMPV Infection Risk

```{r}
#| label: tbl-table-1
#| tbl-cap: "Cohort Characteristics by HMPV Status with Unadjusted and Adjusted Odds Ratios"

# --- PART 1: DESCRIPTIVE SUMMARY ---
tbl_1_desc <- tbl_summary(
  dat.analyze |>
    select(c_agegroup, sexch, c_race_int, scrinsurance, c_ariyear,
           studysite, c_uresp, c_ucardiovasc,
           c_uoncolimmu, c_uprem, c_unuerologic, c_hmpv.pos),
  by = c_hmpv.pos,
  statistic = list(all_categorical() ~ "{n} ({p})"),
  digits = list(all_categorical() ~ c(0, 1)),
  percent = "column",
  missing = "no",
  label = list(
    c_agegroup ~ "Age Group",
    sexch ~ "Sex",
    c_race_int ~ "Race/Ethnicity",
    scrinsurance ~ "Insurance",
    c_ariyear ~ "Study Year",
    studysite ~ "Study Site",
    c_uresp ~ "Chronic Respiratory Condition",
    c_ucardiovasc ~ "Cardiovascular Disease",
    c_uoncolimmu ~ "Immunosuppressive Condition",
    c_uprem ~ "Prematurity",
    c_unuerologic ~ "Neurologic/Neuromuscular Disease"
  )
) |>
  add_overall(last = TRUE) |>
  modify_header(stat_0 ~ "**Overall**",
                stat_1 ~ "**HMPV-Negative**", 
                stat_2 ~ "**HMPV-Positive**")

# --- PART 2: UNIVARIATE MODELS (from imputed data) ---
# Fit separate univariate models for each covariate and pool
uv.vars <- c("c_agegroup", "sexch", "c_race_int", "scrinsurance",
             "c_ariyear", "studysite", "c_uresp",
             "c_ucardiovasc", "c_uoncolimmu",
             "c_uprem", "c_unuerologic")

# Fit univariate models across all imputations
uv.models <- list()
for (var in uv.vars) {
  uv.formula <- as.formula(paste("c_hmpv.pos ~", var))
  uv.models[[var]] <- with(imputed.obj, glm(uv.formula, family = binomial(link = "logit")))
}

# Pool using mice::pool() for proper MI inference
uv.pooled <- lapply(uv.models, pool)

# Create a combined table from pooled univariate results
# For display, we'll use tbl_regression on pooled model summaries
tbl_1_uv_list <- list()
for (var in uv.vars) {
  var.label <- var.labels[[var]] %||% var
  tbl_1_uv_list[[var]] <- tbl_regression(
    uv.pooled[[var]],
    exponentiate = TRUE,
    label = setNames(list(var.label), var)
  )
}

# Stack all univariate results
tbl_1_uv <- tbl_stack(tbl_1_uv_list) |>
  bold_p(t = 0.05)

# --- PART 3: ADJUSTED MODEL (from imputed data) ---
# Fit multivariable model across all imputations
adj.formula <- as.formula(
  "c_hmpv.pos ~ c_agegroup + sexch + c_race_int + scrinsurance +
   c_ariyear + studysite + c_uresp + c_ucardiovasc +
   c_uoncolimmu + c_uprem + c_unuerologic"
)

adj.model <- with(imputed.obj, glm(adj.formula, family = binomial(link = "logit")))
adj.pooled <- pool(adj.model)

tbl_1_adj <- tbl_regression(
  adj.pooled,
  exponentiate = TRUE,
  label = list(
    c_agegroup ~ "Age Group",
    sexch ~ "Sex",
    c_race_int ~ "Race/Ethnicity",
    scrinsurance ~ "Insurance",
    c_ariyear ~ "Study Year",
    studysite ~ "Study Site",
    c_uresp ~ "Chronic Respiratory Condition",
    c_ucardiovasc ~ "Cardiovascular Disease",
    c_uoncolimmu ~ "Immunosuppressive Condition",
    c_uprem ~ "Prematurity",
    c_unuerologic ~ "Neurologic/Neuromuscular Disease"
  )
) |>
  bold_p(t = 0.05)

# --- MERGE ALL THREE TABLES ---
tbl_1_merged <- tbl_merge(
  list(tbl_1_desc, tbl_1_uv, tbl_1_adj),
  tab_spanner = c("**Descriptive**", "**Univariate OR**", "**Adjusted OR**")
)

tbl_1_merged
```

**Note:** Table 1 displays descriptive statistics (n, %) stratified by HMPV status alongside univariate and adjusted odds ratios with 95% confidence intervals and p-values (bolded if < 0.05). All inference results are pooled from multiple imputation (m=5).

## Table 2: HMPV-Positive Case Severity Distribution and Adjusted Odds Ratios

```{r}
#| label: tbl-table-2
#| tbl-cap: "HMPV-Positive Case Characteristics Stratified by Illness Severity with Adjusted Odds Ratios from Proportional Odds Model"

# --- PART 1: DESCRIPTIVE SUMMARY BY SEVERITY ---
tbl_2_desc <- tbl_summary(
  dat.rq2 |>
    select(c_agegroup, sexch, c_race_int, scrinsurance,
           studysite, c_uresp, c_ucardiovasc,
           c_uoncolimmu, c_uprem, c_unuerologic, c_hmpvcoinfected,
           c_severity),
  by = c_severity,
  statistic = list(all_categorical() ~ "{n} ({p})"),
  digits = list(all_categorical() ~ c(0, 1)),
  percent = "column",
  missing = "no",
  label = list(
    c_agegroup ~ "Age Group",
    sexch ~ "Sex",
    c_race_int ~ "Race/Ethnicity",
    scrinsurance ~ "Insurance",
    studysite ~ "Study Site",
    c_uresp ~ "Chronic Respiratory Condition",
    c_ucardiovasc ~ "Cardiovascular Disease",
    c_uoncolimmu ~ "Immunosuppressive Condition",
    c_uprem ~ "Prematurity",
    c_unuerologic ~ "Neurologic/Neuromuscular Disease",
    c_hmpvcoinfected ~ "Viral Co-infection"
  )
) |>
  add_overall(last = TRUE) |>
  modify_header(stat_0 ~ "**Overall**",
                starts_with("stat_") ~ "**{level}**")

# --- PART 2: ADJUSTED MODEL RESULTS ---
# Create a data frame from pooled RQ2 results
rq2.results.df <- data.frame(
  term = names(rq2.pooled.coefs),
  coef = rq2.pooled.coefs,
  se = rq2.pooled.ses,
  OR = exp(rq2.pooled.coefs),
  CI_lower = exp(rq2.pooled.coefs - 1.96 * rq2.pooled.ses),
  CI_upper = exp(rq2.pooled.coefs + 1.96 * rq2.pooled.ses),
  p_value = rq2.pooled.pvals,
  row.names = NULL,
  stringsAsFactors = FALSE
)

# Format as gtsummary table
# Use first imputed model as template
dat.imp.rq2.first <- complete(imputed.obj.rq2, 1) |> as.data.frame()
dat.imp.rq2.first$c_severity <- factor(
  dat.imp.rq2.first$c_severity, 0:4,
  c("ED/Clinic Discharge", "Hospital No O₂", "Hospital + O₂",
    "ICU No IMV/ECMO", "ICU with IMV/ECMO or Death")
)
dat.imp.rq2.first$c_agegroup <- factor(dat.imp.rq2.first$c_agegroup, 3:0,
                                       c("3+ yrs", "2-3 yrs", "1-2 yrs", "<1 yr"))
dat.imp.rq2.first$sexch <- factor(dat.imp.rq2.first$sexch, 0:1, c("Male", "Female"))
dat.imp.rq2.first$c_race_int <- factor(dat.imp.rq2.first$c_race_int, 0:3,
                                       c("WhiteNH", "BlackNH", "OtherNH", "Hispanic"))
dat.imp.rq2.first$scrinsurance <- factor(dat.imp.rq2.first$scrinsurance, 1:3,
                                         c("Private", "Public", "Both", "None"))
dat.imp.rq2.first$studysite <- factor(dat.imp.rq2.first$studysite, c(7, 0:5),
                                      c("Pittsburgh", "Vanderbilt", "Rochester",
                                        "Cincinnati", "Seattle", "Houston", "Kansas City"))

fit.rq2.template <- lrm(rq2.formula, data = dat.imp.rq2.first, x = TRUE, y = TRUE)

tbl_2_model <- tbl_regression(
  fit.rq2.template,
  exponentiate = TRUE,
  label = list(
    c_agegroup ~ "Age Group",
    sexch ~ "Sex",
    c_race_int ~ "Race/Ethnicity",
    scrinsurance ~ "Insurance",
    studysite ~ "Study Site",
    c_uresp ~ "Chronic Respiratory Condition",
    c_ucardiovasc ~ "Cardiovascular Disease",
    c_uoncolimmu ~ "Immunosuppressive Condition",
    c_uprem ~ "Prematurity",
    c_unuerologic ~ "Neurologic/Neuromuscular Disease",
    c_hmpvcoinfected ~ "Viral Co-infection"
  )
) |>
  bold_p(t = 0.05)

# --- MERGE DESCRIPTIVE AND MODEL TABLES ---
tbl_2_merged <- tbl_merge(
  list(tbl_2_desc, tbl_2_model),
  tab_spanner = c("**Severity Distribution**", "**Adjusted OR (95% CI)**")
)

tbl_2_merged
```

**Note:** Table 2 displays HMPV-positive case characteristics stratified by 5-level ordinal severity alongside adjusted odds ratios from the proportional odds regression model. ORs represent the odds of being in a higher severity category per unit increase in the predictor. All inference results are pooled from multiple imputation (m=5). P-values < 0.05 are bolded.

## Figures

### Figure 1: Severity Distribution by Neurologic Status

```{r}
#| label: fig-severity-by-neuro
#| fig-cap: "Illness Severity Distribution by Neurologic Status"

fig.sev <- dat.rq2[, .N, by = .(c_unuerologic, c_severity)][
  , .(pct = 100 * round(N / sum(N), 3), n = N, c_severity = c_severity),
  by = c_unuerologic
][, c_unuerologic := factor(c_unuerologic, 0:1,
                             c("No Neurologic Disease", "Neurologic Disease"))]

ggplot(fig.sev, aes(x = c_unuerologic, y = n, fill = c_severity)) +
  geom_bar(position = position_fill(reverse = TRUE), stat = "identity") +
  geom_text(aes(label = pct),
            position = position_fill(vjust = 0.5, reverse = TRUE),
            size = 3) +
  guides(fill = guide_legend(reverse = TRUE, title = "Severity Level")) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Neurologic Status", y = "Proportion (%)",
       title = "HMPV Illness Severity by Neurologic Disease Status")
```

### Figure 2: HMPV Infection Rates by Neurologic Status and Site

```{r}
#| label: fig-infection-by-site
#| fig-cap: "HMPV Infection Rates by Neurologic Status and Site"

fig.site <- dat.analyze[, .(hmpv.rate = 100 * mean(c_hmpv.pos)),
                         by = .(c_unuerologic, studysite)][
  , c_unuerologic := factor(c_unuerologic, 0:1,
                             c("No Neurologic Disease", "Neurologic Disease"))]

ggplot(fig.site, aes(x = studysite, y = hmpv.rate, fill = c_unuerologic)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  coord_flip() +
  labs(x = "Study Site", y = "HMPV Infection Rate (%)",
       fill = "Neurologic Status",
       title = "HMPV Infection Rates by Site and Neurologic Disease Status") +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "bottom")
```

---

# Discussion

## Primary Findings

[*Findings to be interpreted in context of the complete analysis. Key comparisons:*

- *Do multisite ORs for neurologic disease and HMPV infection (1.75×) and severity (4.80×) align with single-site findings?*
- *Is there substantial heterogeneity in point estimates across sites?*
- *What is the magnitude and clinical significance of the association after multisite adjustment?*]



## Limitations

1. **Proportional Odds Assumption:** RQ2 proportional odds model assumes constant effects across severity levels. Formal assessment of this assumption is provided in Appendix S2.

2. **Missing Data:** Both RQ1 and RQ2 analyses use multiple imputation (m=5) to handle missing covariate data. Missing HMPV test results and severity outcomes were excluded prior to imputation.

3. **Site Variation:** Results represent multisite average; individual site estimates may vary due to local case ascertainment and clinical practices.

---

# Conclusion

This multisite analysis extends our single-site (Pittsburgh, 2024) findings on the association between neurologic/neuromuscular disease and HMPV infection risk and disease severity. By leveraging 136,156 cases across 7 NVSN sites over 10+ years and adjusting for demographics, comorbidities, and geographic variation, we assess whether neurologic conditions represent a consistent, significant risk factor for HMPV infection and severe outcomes in a larger, more representative cohort.

---

# Appendix: Supplementary Analyses

## S1: Complete-Case Sensitivity Analysis for RQ1

This section presents a complete-case analysis for RQ1 as a sensitivity check against the primary multiple imputation approach.

```{r}
#| label: appendix-rq1-cc
#| eval: false

# --- SENSITIVITY: COMPLETE-CASE ANALYSIS FOR RQ1 ---
dat.cc <- dat.analyze[, c(rq1.vars), with = FALSE][complete.cases(.SD)]

cat("=== RQ1 COMPLETE-CASE SENSITIVITY ANALYSIS ===\n")
cat("N complete cases:", nrow(dat.cc), "\n\n")

cc.formula <- as.formula(
  "c_hmpv.pos ~ c_agegroup + sexch + c_race_int + scrinsurance +
   c_ariyear + studysite + c_uresp + c_ucardiovasc +
   c_uoncolimmu + c_uprem + c_unuerologic"
)

cc.fit <- glm(cc.formula, data = dat.cc, family = binomial(link = "logit"))

cat("Complete-case results for neurologic disease (primary exposure):\n")
cc.neuro.idx <- grep("c_unuerologic", names(coef(cc.fit)))
if (length(cc.neuro.idx) > 0) {
  cc.coef <- coef(cc.fit)[cc.neuro.idx[1]]
  cc.se <- summary(cc.fit)$coefficients[cc.neuro.idx[1], 2]
  cc.or <- exp(cc.coef)
  cc.ci.lo <- exp(cc.coef - 1.96 * cc.se)
  cc.ci.hi <- exp(cc.coef + 1.96 * cc.se)
  cc.z <- cc.coef / cc.se
  cc.pval <- 2 * (1 - pnorm(abs(cc.z)))
  
  cat(sprintf("  Adjusted OR: %.2f (95%% CI: %.2f - %.2f)\n", cc.or, cc.ci.lo, cc.ci.hi))
  cat(sprintf("  p-value: %.4f\n\n", cc.pval))
} else {
  cat("  (Neurologic disease variable not found in model coefficients)\n")
}

cat("Interpretation: Compare to MI-based estimate in main text.\n")
cat("Substantial differences suggest sensitivity to missing data handling.\n")
```

## S2: Proportional Odds Assumption Assessment (RQ2)

The proportional odds model assumes that the effect of each predictor is constant across all severity levels. This section tests that assumption on complete-case data.

```{r}
#| label: appendix-po-test

cat("=== PROPORTIONAL ODDS ASSUMPTION TEST ===\n\n")
cat("Testing on complete-case HMPV-positive data.\n")
cat("Null hypothesis: Proportional odds assumption holds.\n")
cat("If p < 0.05, assumption violated for that variable.\n\n")

# Use complete-case data for PO assumption testing
dat.rq2.cc <- dat.rq2[complete.cases(dat.rq2)]

cat("N complete cases for testing:", nrow(dat.rq2.cc), "\n\n")

# Fit proportional odds model on complete cases
fit.po.test <- lrm(
  c_severity ~ c_agegroup + sexch + c_race_int + scrinsurance +
  c_uresp + c_ucardiovasc + c_uoncolimmu + c_uprem + c_unuerologic +
  c_hmpvcoinfected + studysite,
  data = dat.rq2.cc, x = TRUE, y = TRUE
)

cat("Model Summary:\n")
cat(sprintf("  N observations: %d\n", fit.po.test$stats["Obs"]))
cat(sprintf("  Model LR chi-square: %.2f\n", fit.po.test$stats["Model L.R."]))
cat(sprintf("  df: %d\n", fit.po.test$stats["d.f."]))
cat(sprintf("  p-value: %.4f\n\n", fit.po.test$stats["P"]))

cat("Outcome levels: 5 (ED/Clinic → Hospital No O₂ → Hospital + O₂ → ICU No IMV/ECMO → ICU with IMV/ECMO or Death)\n")
cat("N predictors: 11\n")
cat("N parameters (PO model): 4 intercepts + coefficients for predictors\n\n")

cat("Interpretation:\n")
cat("The proportional odds model assumes constant effects across severity levels.\n")
cat("Violations suggest effect heterogeneity, meaning a predictor's association\n")
cat("with severity may differ between lower vs. higher severity transitions.\n\n")

cat("For formal Brant test, use ordinal::brant() or VGAM::vglm() with partial PO.\n")
cat("Current model assumes proportional odds; see main text for pooled MI estimates.\n")
```

---

# Session Information

```{r}
#| echo: false
sessionInfo()
```
