---
title: "Neurologic Disease and HMPV Infection Risk: NVSN Multisite Analysis"
subtitle: "2016-2026, Complete Case Analysis"
format:
  html:
    output-file: "HMPV_Neurology_Analysis.html"
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
    code-fold: true
execute:
  warning: false
fig-cap-location: top
abstract-title: "Overview"
abstract: "This document presents a complete case analysis investigating
  whether children with neurologic/neuromuscular disease experience increased
  risk of HMPV (human metapneumovirus) infection and disease severity across
  7 NVSN sites from 2016-2026. We compare findings to a previous single-site
  (Pittsburgh) analysis that found neurologic disease associated with 1.75×
  higher odds of HMPV infection and 4.80× higher odds of severe illness."
---

```{r setup}
library(data.table)
library(gtsummary)
library(ggplot2)
library(ggpubr)
library(MASS)
library(brant)
source("wald_function.R")
```

```{r data}
dat <- fread("Data/Pitt_Anna_HMPV_SEP26.csv", na.strings = "")
```

# Data Preparation

```{r exposure_variable}
# Neurologic/neuromuscular disease status (explicit NA handling)
dat[, neuro.status := fcase(
  c_unuerologic == 1, "Neurologic Disease",
  c_unuerologic == 0, "No Neurologic Disease")]
dat[, neuro.status := factor(
  neuro.status,
  c("No Neurologic Disease", "Neurologic Disease"))]
```

```{r outcome_variables}
# Primary outcome: HMPV infection status
dat[, hmpv.pos := tmpv == 1]
dat[, hmpv.status := factor(
  hmpv.pos, c(TRUE, FALSE), c("HMPV-Pos", "HMPV-neg"))]

# Secondary outcome: Illness severity (5-level ordinal)
# First, create helper variables for O2/respiratory support
dat[c_finalstatus == 1, c_hospO2 := fcase(
  c_suppoxy == 1, TRUE, c_blowby==1, TRUE, c_cpap==1, TRUE,
  inptICU == 1, TRUE,
  c_intubated == 1, TRUE,
  inptEcmo == 1, TRUE,
  default = FALSE)]  # If admitted but no ICU/intubation/ECMO markers

dat[, c_mv_ecmo := fcase(
  c_intubated == 1, TRUE,
  inptEcmo == 1, TRUE,
  default = FALSE)]

dat[, c_icu := inptICU == 1]

# Construct 5-level severity outcome
dat[, severity := fcase(
  c_finalstatus == 2, 1L,  # ED discharge
  c_finalstatus == 1 & !c_hospO2, 2L,  # Admitted, no support
  c_finalstatus == 1 & c_hospO2 & !c_icu, 3L,  # Admitted with O2/resp support
  c_finalstatus == 1 & c_icu & !c_mv_ecmo, 4L,  # ICU admission
  c_finalstatus == 1 & c_mv_ecmo, 5L)]  # MV/ECMO

dat[, severity := factor(
  severity, 1:5,
  c("ED discharge", "Admitted - no support", "Admitted - O2/resp support",
    "ICU admission", "Mechanical ventilation/ECMO"))]
```

```{r covariate_encoding}
# Site
dat[, studysite := factor(
  studysite, c(8, 1:6),
  c("Pittsburgh", "Nashville", "Rochester", "Cincinnati", "Seattle", "Houston",
    "Kansas City"))]

# Demographics
dat[, sexch := factor(sexch, 2:1, c("Female", "Male"))]
dat[, c_race_int := factor(
  c_race_int, c(1, 2, 4, 3, 8),
  c("White NH", "Black NH", "Hispanic", "Other NH", "Unknown"))]

# Insurance (verify levels in data)
dat[, insurance.cat := factor(
  scrinsurance, c(1, 2, 3, 4, 8),
  c("Private", "Public", "Both Public/Private", "None", "Unknown"))]

# Season - Exclude Pilot year
dat[, c_ariyear := factor(
  c_ariyear,
  c(2:11),
  c("2016-17", "2017-18", "2018-19", "2019-20", "2020-21",
    "2021-22", "2022-23", "2023-24", "2024-25", "2025-26"))]

# Comorbidities - explicit TRUE/FALSE/NA encoding
dat[, respiratory.cond := fcase(
  c_uresp == 1, TRUE,
  c_uresp == 0, FALSE)]

dat[, cardiovasc.cond := fcase(
  c_ucardiovasc == 1, TRUE,
  c_ucardiovasc == 0, FALSE)]

dat[, immunocomp.cond := fcase(
  c_uoncolimmu == 1, TRUE,
  c_uoncolimmu == 0, FALSE)]

# Prematurity - default assume full term
dat[, prematurity.cond := fcase(
  c_uprem == 1, TRUE,
  default=FALSE)]
```

```{r complete_case_restriction}
# Define required variables
required.vars <- c(
  "hmpv.pos", "neuro.status", "c_ageyears", "sexch", "c_race_int",
  "insurance.cat", "respiratory.cond", "cardiovasc.cond",
  "immunocomp.cond", "prematurity.cond", "studysite", "c_ariyear")

# Create complete case indicator
dat[, complete.case := complete.cases(.SD), .SDcols = required.vars]

# Document exclusions
exclusion.summary <- dat[, .(
  total = .N,
  missing_hmpv = sum(is.na(hmpv.pos)),
  missing_neuro = sum(is.na(neuro.status)),
  missing_age = sum(is.na(c_ageyears)),
  missing_sex = sum(is.na(sexch)),
  missing_race = sum(is.na(c_race_int)),
  missing_insurance = sum(is.na(insurance.cat)),
  missing_resp = sum(is.na(respiratory.cond)),
  missing_cardio = sum(is.na(cardiovasc.cond)),
  missing_immuno = sum(is.na(immunocomp.cond)),
  missing_prem = sum(is.na(prematurity.cond)),
  missing_site = sum(is.na(studysite)),
  missing_year = sum(is.na(c_ariyear)),
  complete_cases = sum(complete.case))]

# Subset to complete cases
inc <- dat[complete.case == TRUE]
```

# Study Population

[@tbl-exclusions summarizes case inclusion. Starting with
`r format(nrow(dat), big.mark = ",")` total ARI cases, we retained
`r format(nrow(inc), big.mark = ",")` cases with complete data on HMPV test
results, neurologic status, demographics, and comorbidities.]{.aside}

```{r}
#| label: tbl-exclusions
#| tbl-cap: "Case inclusion summary"
#| echo: false
exclusion.columns <- names(exclusion.summary)[exclusion.summary > 0]
knitr::kable(
  exclusion.summary[, ..exclusion.columns],
  format.args = list(big.mark = ","))
```

# Primary Analysis: HMPV Infection Risk

```{r infection_models}
# Unadjusted model
fit.unadj <- glm(
  hmpv.pos ~ neuro.status,
  data = inc,
  family = binomial)

# Adjusted model
fit.adj <- glm(
  hmpv.pos ~ neuro.status + c_ageyears + sexch + c_race_int +
    insurance.cat + respiratory.cond + cardiovasc.cond +
    immunocomp.cond + prematurity.cond + studysite + c_ariyear,
  data = inc,
  family = binomial)
```

```{r infection_tables}
# Descriptive statistics
desc.labs <- list(
  neuro.status = "Neurologic/neuromuscular disease",
  c_ageyears = "Age (years)",
  sexch = "Sex",
  c_race_int = "Race/ethnicity",
  insurance.cat = "Insurance type",
  respiratory.cond = "Respiratory condition",
  cardiovasc.cond = "Cardiovascular condition",
  immunocomp.cond = "Immunocompromised",
  prematurity.cond = "Prematurity",
  studysite = "Study site",
  c_ariyear = "Season")

tbl.desc <- tbl_summary(
  inc,
  by = hmpv.status,
  include = names(desc.labs),
  label = desc.labs,
  missing = "no") |>
  add_overall(last = TRUE)

# Unadjusted ORs
tbl.unadj <- tbl_regression(
  fit.unadj,
  tidy_fun = tidy.glm.wald,
  exponentiate = TRUE,
  include = neuro.status,
  label = desc.labs) |>
  bold_p() |>
  modify_header(estimate = "**OR**")

# Adjusted ORs
tbl.adj <- tbl_regression(
  fit.adj,
  label = desc.labs,
  tidy_fun = tidy.glm.wald,
  exponentiate = TRUE,
  show_single_row = c(respiratory.cond, cardiovasc.cond,
                      immunocomp.cond, prematurity.cond)) |>
  bold_p() |>
  modify_header(estimate = "**AOR**") |>
  remove_abbreviation() |>
  modify_abbreviation("AOR = Adjusted Odds Ratio")
```

[@tbl-infection-risk presents HMPV infection risk analysis. The left panel
shows descriptive statistics stratified by HMPV status. The middle panel
shows unadjusted odds ratios, and the right panel shows adjusted odds ratios
controlling for demographics, comorbidities, site, and season.]{.aside}

```{r}
#| label: tbl-infection-risk
#| tbl-cap: "HMPV infection risk by neurologic status"
#| echo: false
tbl_merge(
  list(tbl.desc, tbl.unadj, tbl.adj),
  tab_spanner = c("Descriptive", "Unadjusted", "Adjusted")) |>
  remove_footnote_header() |>
  modify_header(label = "**Characteristic**")
```

# Site-Level Variation

```{r fig_infection_by_site}
# Calculate HMPV positivity rates by site, neurologic status, and CoAg period
inc[, period := fcase(
  c_ariyear %in% c("2016-17", "2017-18", "2018-19", "2019-20",
                   "2020-21"), "CoAg1",
  default = "CoAg2")]
rates.by.site <- inc[, .(
  hmpv.rate = mean(hmpv.pos, na.rm = TRUE),
  n.pos = sum(hmpv.pos, na.rm = TRUE),
  n.total = .N),
  by = .(neuro.status, studysite, period)]

# Add 95% CI using normal approximation
rates.by.site[, se := sqrt(hmpv.rate * (1 - hmpv.rate) / n.total)]
rates.by.site[, lwr := pmax(0, hmpv.rate - 1.96 * se)]
rates.by.site[, upr := pmin(1, hmpv.rate + 1.96 * se)]
```

[@fig-infection-rates shows HMPV positivity rates by coag period, study site and
neurologic status. Error bars represent 95% confidence intervals.]{.aside}

```{r}
#| label: fig-infection-rates
#| fig-cap: "HMPV infection rates by site and neurologic status"
#| fig-width: 9
#| fig-height: 5
#| echo: false
p1 <- ggplot(
  rates.by.site,
  aes(x = studysite, y = hmpv.rate, color = neuro.status,
      group = neuro.status)) +
  facet_wrap(~period) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_errorbar(
    aes(ymin = lwr, ymax = upr),
    width = 0.2, position = position_dodge(width = 0.4)) +
  geom_line(position = position_dodge(width = 0.4)) +
  labs(x = "", y = "HMPV Positivity Rate",
       color = "Neurologic Status") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_pubclean() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")
p1
```

# Secondary Analysis: Disease Severity Among HMPV+ Cases

```{r severity_analysis}
# Subset to HMPV-positive cases
hmpv.cases <- inc[hmpv.pos == TRUE & !is.na(severity)]

# Descriptive summary by severity level
tbl.sev.desc <- tbl_summary(
  hmpv.cases,
  by = severity,
  include = c(neuro.status, c_ageyears, sexch, c_race_int, insurance.cat,
              respiratory.cond, cardiovasc.cond, immunocomp.cond,
              prematurity.cond, studysite, c_ariyear),
  label = desc.labs,
  missing = "no")

# Proportional odds model
fit.severity <- polr(
  severity ~ neuro.status + c_ageyears + sexch + c_race_int +
    insurance.cat + respiratory.cond + cardiovasc.cond +
    immunocomp.cond + prematurity.cond + studysite + c_ariyear,
  data = hmpv.cases,
  Hess = TRUE)

tbl.sev.adj <- tbl_regression(
  fit.severity,
  label = desc.labs,
  tidy_fun = tidy.glm.wald,
  exponentiate = TRUE,
  show_single_row = c(respiratory.cond, cardiovasc.cond,
                      immunocomp.cond, prematurity.cond)) |>
  bold_p() |>
  modify_header(estimate = "**OR**") |>
  modify_footnote(
    estimate ~ "Proportional odds ratio for elevated severity")
```

[@tbl-severity-analysis presents illness severity among HMPV-positive cases.
The proportional odds model estimates the odds of being in a higher severity
category. The proportional odds assumption was tested using the Brant test;
see @tbl-brant-test for results.]{.aside}

```{r}
#| label: tbl-severity-analysis
#| tbl-cap: "Illness severity among HMPV-positive cases"
#| echo: false
tbl_merge(
  list(tbl.sev.desc, tbl.sev.adj),
  tab_spanner = c("Descriptive", "Adjusted OR")) |>
  remove_footnote_header() |>
  modify_header(label = "**Characteristic**")
```

## Proportional Odds Assumption

The omnibus test (p < 0.001) indicates violation of the proportional odds assumption. 

```{r}
# Test proportional odds assumption
brant.test <- brant(fit.severity)
```

The Brant test evaluates whether the proportional odds assumption holds—that is, whether the effect of each covariate is the same across all severity thresholds. A non-significant p-value (p > 0.05) indicates the assumption is satisfied for that variable.

```{r}
#| label: tbl-brant-test
#| tbl-cap: "Brant test for proportional odds assumption"
#| echo: false
brant.results <- data.frame(
  Variable = rownames(brant.test),
  Chi_Square = brant.test[, "X2"],
  DF = brant.test[, "df"],
  P_Value = brant.test[, "probability"])

knitr::kable(
  brant.results,
  digits = c(0, 2, 0, 3),
  col.names = c("Variable", "χ²", "df", "p-value"),
  row.names = FALSE)
```

Key violations include:

- **Neurologic status** (p = 0.014): The primary exposure shows non-proportional effects across severity levels
- **Site effects**: Multiple sites (Kansas City, Houston, Seattle, Nashville) show significant violations
- **Comorbidities**: Immunocompromised status (p < 0.001) and prematurity (p < 0.001) violate the assumption
- **Temporal effects**: Several recent seasons (2021-26) show non-proportional effects

[@fig-brant-pvalues shows p-values from the Brant test for each covariate, with the red dashed line indicating the α = 0.05 threshold.]{.aside}

```{r}
#| label: fig-brant-pvalues
#| fig-cap: "Brant test p-values by covariate"
#| fig-width: 8
#| fig-height: 7
#| echo: false
# Remove Omnibus row for plotting
brant.plot <- brant.results[brant.results$Variable != "Omnibus", ]

# Clean variable names for display
brant.plot$Variable.clean <- gsub("neuro.statusNeurologic Disease",
                                  "Neurologic Disease*",
                                  brant.plot$Variable)
brant.plot$Variable.clean <- gsub("c_ageyears", "Age", 
                                  brant.plot$Variable.clean)
brant.plot$Variable.clean <- gsub("sexchMale", "Sex: Male", 
                                  brant.plot$Variable.clean)
brant.plot$Variable.clean <- gsub("c_race_int", "Race: ", 
                                  brant.plot$Variable.clean)
brant.plot$Variable.clean <- gsub("insurance.cat", "Insurance: ", 
                                  brant.plot$Variable.clean)
brant.plot$Variable.clean <- gsub("TRUE", "", 
                                  brant.plot$Variable.clean)
brant.plot$Variable.clean <- gsub("\\.cond", "", 
                                  brant.plot$Variable.clean)
brant.plot$Variable.clean <- gsub("studysite", "Site: ", 
                                  brant.plot$Variable.clean)
brant.plot$Variable.clean <- gsub("c_ariyear", "Season: ", 
                                  brant.plot$Variable.clean)

# Highlight primary exposure
brant.plot$is.primary <- grepl("\\*$", brant.plot$Variable.clean)

# Order by p-value
brant.plot <- brant.plot[order(-brant.plot$P_Value), ]
brant.plot$Variable.clean <- factor(brant.plot$Variable.clean,
                                    levels = brant.plot$Variable.clean)

ggplot(brant.plot, aes(x = P_Value, y = Variable.clean,
                       color = is.primary)) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("FALSE" = "gray40", "TRUE" = "blue"),
                     guide = "none") +
  scale_x_continuous(trans = "log10", breaks = c(0.001, 0.01, 0.05, 0.1, 1)) +
  labs(x = "p-value (log scale)", y = "") +
  theme_pubclean() +
  theme(axis.text.y = element_text(size = 9))
```

**Interpretation:** While the proportional odds model provides a parsimonious summary, the violations suggest neurologic disease may have differential effects at different severity thresholds. The binary severity sensitivity analysis (below) provides a complementary approach less sensitive to these violations.

# Severity Distribution

```{r fig_severity_dist}
# Calculate severity proportions
sev.dist <- hmpv.cases[, .N, by = .(neuro.status, severity)]
sev.dist[, prop := N / sum(N), by = neuro.status]
```

[@fig-severity-distribution shows the distribution of illness severity among
HMPV-positive cases, stratified by neurologic status.]{.aside}

```{r}
#| label: fig-severity-distribution
#| fig-cap: "Severity distribution among HMPV-positive cases"
#| fig-width: 8
#| fig-height: 5
#| echo: false
p2 <- ggplot(sev.dist, aes(x = neuro.status, y = prop, fill = severity)) +
  geom_col(position = "fill") +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "", y = "Proportion of HMPV+ Cases",
       fill = "Illness Severity") +
  theme_pubclean() +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.position = "bottom") +
  guides(fill = guide_legend(ncol = 2))
p2
```

# Comparison to Pittsburgh 2016-21 Findings

**Previous single-site findings (Pittsburgh only):**

- HMPV infection risk: OR = 1.75 (95% CI: 1.19, 2.56)
- HMPV severity: OR = 4.80 (95% CI: 2.01, 11.47)

**Current multisite findings:**
```{r comparison_summary}
# Extract ORs from models
infection.or <- exp(coef(fit.adj)["neuro.statusNeurologic Disease"])
infection.ci <- exp(confint.default(
  fit.adj, "neuro.statusNeurologic Disease"))

severity.or <- exp(coef(fit.severity)["neuro.statusNeurologic Disease"])
severity.ci <- exp(confint.default(
  fit.severity, "neuro.statusNeurologic Disease"))

comparison <- data.frame(
  Outcome = c("HMPV Infection Risk", "HMPV Severity (among HMPV+)"),
  OR = c(infection.or, severity.or),
  CI_Lower = c(infection.ci[1], severity.ci[1]),
  CI_Upper = c(infection.ci[2], severity.ci[2]))
```

```{r}
#| echo: false
knitr::kable(
  comparison,
  digits = 2,
  col.names = c("Outcome", "AOR", "95% CI Lower", "95% CI Upper"))
```

# Sensitivity Analyses {.appendix}

## Model Without Site/Year Adjustment

```{r sensitivity_no_site}
fit.no.site <- glm(
  hmpv.pos ~ neuro.status + c_ageyears + sexch + c_race_int +
    insurance.cat + respiratory.cond + cardiovasc.cond +
    immunocomp.cond + prematurity.cond,
  data = inc, family = binomial)

tbl.no.site <- tbl_regression(
  fit.no.site,
  label = desc.labs,
  tidy_fun = tidy.glm.wald,
  exponentiate = TRUE,
  include = neuro.status) |>
  bold_p() |>
  modify_header(estimate = "**AOR**")
```

```{r}
#| echo: false
tbl.no.site
```

## Dichotomous Severity Outcome

```{r sensitivity_binary_severity}
# Define severe illness as ICU admission or MV/ECMO
hmpv.cases[, severe := severity %in% c("ICU admission",
                                       "Mechanical ventilation/ECMO")]

fit.severe.binary <- glm(
  severe ~ neuro.status + c_ageyears + sexch + c_race_int +
    insurance.cat + respiratory.cond + cardiovasc.cond +
    immunocomp.cond + prematurity.cond + studysite + c_ariyear,
  data = hmpv.cases,
  family = binomial)

tbl.severe.binary <- tbl_regression(
  fit.severe.binary,
  label = desc.labs,
  tidy_fun = tidy.glm.wald,
  exponentiate = TRUE,
  include = neuro.status) |>
  bold_p() |>
  modify_header(estimate = "**AOR**") |>
  modify_footnote(
    estimate ~ "Adjusted odds ratio for severe illness (ICU/MV/ECMO)")
```

```{r}
#| echo: false
tbl.severe.binary
```
